---
title: "Anaylises for SONATA review"
subtitle: "Purpose field"
author: "Alicia Valdés"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
  html_notebook: default
---

This is a script to perform analyses for the SONATA review paper.

# Load libraries

```{r}
library(here)
library(readxl)
library(tidyverse)
```

# Read data

In this data, in the sheet "env" I have edited the environmental variables manually in order to ALWAYS have ";" separating the different variables in each column. 

```{r}
sonata_data <-read_excel(here("data", "edited","database_sonata_v2.2_AV.xlsx"),
                         sheet = "env")
sonata_data
```

# Check for duplicate IDs

```{r}
sonata_data %>% count(ID) %>% filter(n > 1)
```

```{r}
sonata_data %>% mutate(row_number = row_number()) %>% filter(ID == 2495)
```

```{r}
# Remove row number of the wrong entry
sonata_data <- sonata_data %>% slice(-619)
```

# Clean environmental variables

## Convert all to 0/1

```{r}
sonata_data <- sonata_data %>%
  mutate(Climate_bin = if_else(is.na(Climate) | Climate == "", 0, 
                               if_else(Climate == "None", 0, 1)),
         Soil_bin = if_else(is.na(Soil) | Soil == "", 0, 
                            if_else(Soil == "None" | Soil == "Soil: none", 0,
                                    1)),
         Topography_bin = if_else(is.na(Topography) | Topography == "", 0, 
                                  if_else(Topography == "None" | 
                                            Topography == "None specified" |
                                            Topography == "Not explicitly", 0,
                                          1)),
         Biotic_bin = if_else(is.na(Biotic) | Biotic == "", 0, 
                              if_else(Biotic == "None", 0, 1)))
```

## Climate

```{r}
sonata_climate <- sonata_data %>%
  select(ID, starts_with("Climate")) %>%
  separate(Climate, 
           # There is at most 6 different climatic variables
           # Separate into 6 cols
           into = paste0("Climate", 1:6), 
           # The different variables in each col are always separated by ";"
           sep = ";", 
           fill = "right",
           remove = FALSE, extra = "warn") %>%
  # Remove whitespace
  mutate(across(starts_with("Climate"), str_trim)) %>%
  rowwise() %>%
  # Create new cols to store data on most important Climate variables
  mutate(
    # Any of the Climate cols contains temperature
    Clim_temp = any(str_detect(c_across(Climate1:Climate6),
                               regex("temperature", ignore_case = TRUE))),
    # Any of the Climate cols contains precipitation, rainfall or drought
    Clim_precip = any(str_detect(c_across(Climate1:Climate6),
                                 regex("precipitation|rainfall|drought",
                                       ignore_case = TRUE))) ,
    # Any of the Climate cols contains any word containing "radi" 
    # (like "radiation", "irradiation") and the exact word "PAR"
    Clim_rad = any(str_detect(c_across(Climate1:Climate6), 
                              regex("radi|\\bPAR\\b", ignore_case = TRUE))),
    # Any of the Climate cols contains humidity
    Clim_humid = any(str_detect(c_across(Climate1:Climate6), 
                                regex("humidity", ignore_case = TRUE))),
    # Any of the Climate cols contains wind
    Clim_wind = any(str_detect(c_across(Climate1:Climate6), 
                               regex("wind", ignore_case = TRUE))),
    # Any of the Climate cols contains evapotranspiration
    Clim_evap = any(str_detect(c_across(Climate1:Climate6), 
                               regex("evapotranspiration",
                                     ignore_case = TRUE))),
    # There is sth in any of the Climate cols and
    # all previous variables are FALSE
    Clim_other = if_else(Climate1 != 1 & Climate_bin == 1 & 
                           all(is.na(c_across(Clim_temp:Clim_evap))), TRUE, NA),
    # Climatic variables not specified
    Clim_unspecif = if_else(Climate1 == 1, TRUE, NA)
    ) %>%
  ungroup()
```

## Soil

```{r}
sonata_soil <- sonata_data %>%
  select(ID, starts_with("Soil")) %>%
  separate(Soil, 
           # There is at most 8 different soil variables
           # Separate into 8 cols
           into = paste0("Soil", 1:8), 
           # The different variables in each col are always separated by ";"
           sep = ";", 
           fill = "right",
           remove = FALSE, extra = "warn") %>%
  # Remove whitespace
  mutate(across(starts_with("Soil"), str_trim)) %>%
  rowwise() %>%
  # Create new cols to store data on most important Soil variables
  mutate(
    # Any of the soil cols contains any word containing "type" or "class"
    Soil_type = any(str_detect(c_across(Soil1:Soil8), 
                               regex("type|class", ignore_case = TRUE))),
    # Any of the Soil cols contains texture
    Soil_text = any(str_detect(c_across(Soil1:Soil8), 
                               regex("texture", ignore_case = TRUE))),
    # Any of the soil cols contains moisture, water or wetness
    Soil_moist = any(str_detect(c_across(Soil1:Soil8),
                                regex("moisture|water|wetness",
                                      ignore_case = TRUE))),
    # Any of the soil cols contains depth
    Soil_depth = any(str_detect(c_across(Soil1:Soil8), 
                                regex("depth", ignore_case = TRUE))),
    # Any of the soil cols contains pH or acidity
    Soil_ph = any(str_detect(c_across(Soil1:Soil8), 
                             regex("pH|acidity", ignore_case = TRUE))),
    # Any of the soil cols contains carbon
    Soil_carbon = any(str_detect(c_across(Soil1:Soil8), 
                                 regex("carbon", ignore_case = TRUE))),
    # Any of the soil cols contains roughness
    Soil_rough = any(str_detect(c_across(Soil1:Soil8), 
                                regex("roughness", ignore_case = TRUE))),
    # Any of the soil cols contains any word containing "ferti"
    Soil_ferti = any(str_detect(c_across(Soil1:Soil8), 
                               regex("ferti", ignore_case = TRUE))),
    # Any of the soil cols contains nitrogen
    Soil_nitro = any(str_detect(c_across(Soil1:Soil8), 
                                regex("nitrogen", ignore_case = TRUE))),
    # Any of the soil cols contains bulk
    Soil_bulk = any(str_detect(c_across(Soil1:Soil8), 
                               regex("bulk", ignore_case = TRUE))),
    # There is sth in any of the Soil cols and
    # all previous variables are FALSE
    Soil_other = if_else(Soil1 != 1 & Soil_bin == 1 & 
                           all(is.na(c_across(Soil_type:Soil_bulk))), TRUE, NA),
    # Soil variables not specified
    Soil_unspecif = if_else(Soil == 1, TRUE, NA)
    ) %>%
  ungroup()
```

## Topography

```{r}
sonata_topo <- sonata_data %>%
  select(ID, starts_with("Topo")) %>%
  separate(Topography, 
           # There is at most 7different topo variables
           # Separate into 7 cols
           into = paste0("Topo", 1:7), 
           # The different variables in each col are always separated by ";"
           sep = ";", 
           fill = "right",
           remove = FALSE, extra = "warn") %>%
  # Remove whitespace
  mutate(across(starts_with("Topo"), str_trim)) %>%
  rowwise() %>%
  # Create new cols to store data on most important Topo variables
  mutate(
    # Any of the Topo cols contains aspect
    Topo_aspect = any(str_detect(c_across(Topo1:Topo7), 
                               regex("aspect", ignore_case = TRUE))),
    # Any of the Topo cols contains slope
    Topo_slope = any(str_detect(c_across(Topo1:Topo7), 
                               regex("slope", ignore_case = TRUE))),
    # Any of the Topo cols contains elevation or altitude
    Topo_elev = any(str_detect(c_across(Topo1:Topo7), 
                               regex("elevation|altitude", 
                                     ignore_case = TRUE))),
    # Any of the Topo cols contains TWI or wetness
    Topo_twi = any(str_detect(c_across(Topo1:Topo7), 
                               regex("TWI|wetness", ignore_case = TRUE))),
    # Any of the Topo cols contains curvature
    Topo_curv = any(str_detect(c_across(Topo1:Topo7), 
                               regex("curvature", ignore_case = TRUE))),
    # There is sth in any of the Topo cols and
    # all previous variables are FALSE
    Topo_other = if_else(Topo1 != 1 & Topography_bin == 1 & 
                           all(is.na(c_across(Topo_aspect:Topo_curv))), TRUE,
                         NA),
    # Topo variables not specified
    Topo_unspecif = if_else(Topography == 1, TRUE, NA)
    ) %>%
  ungroup()
```

# Merged data

```{r}
sonata_env_data_clean <- sonata_data %>% 
  # Add Climate
  left_join(sonata_climate %>% select(ID, Clim_temp:Clim_unspecif)) %>%
  # Add Soil
  left_join(sonata_soil %>% select(ID, Soil_type:Soil_unspecif)) %>%
  # Add Topo
  left_join(sonata_topo %>% select(ID, Topo_aspect:Topo_unspecif)) %>%
  mutate(across(Clim_temp:Topo_unspecif, ~ ifelse(is.na(.), 0, 1)))
```

# Save as csv

```{r}
write_csv(sonata_env_data_clean, "data/clean/sonata_env_data_clean.csv")
```


# Session info

```{r}
sessionInfo()
```
