---
title: "Anaylises for SONATA review"
subtitle: "Merge all databases"
author: "Alicia Valdés"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
  html_notebook: default
---

This is a script to merge all databases for the SONATA review paper.

# Load libraries

```{r}
library(here)
library(readxl)
library(tidyverse)
library(stringr)
```

# Read data

## Main DB

```{r}
sonata_main <- read_excel(
  here("data", "edited","to_merge", "database_sonata_v2.2_extent_clean.xlsx")) %>%
  # Keep only fields that have not been cleaned in other DBs
  select(ID, Reader, Year, Purpose, `keyword 1`, `keyword 2`, `keyword 3`,
         Extent, # Needs cleaning!
         Highlight) %>%
  # Rename keyword fields
  rename(keyword1 = `keyword 1`, keyword2 = `keyword 2`, keyword3 = `keyword 3`)
sonata_main
```

Remove duplicate:

```{r}
sonata_main %>% count(ID) %>% filter(n > 1)
```

```{r}
sonata_main %>% mutate(row_number = row_number()) %>% filter(ID == 2495)
```

```{r}
# Remove row number of the wrong entry
sonata_main <- sonata_main %>% slice(-619)
```

## Countries DB

```{r}
sonata_countries <-read_excel(
  here("data", "edited","to_merge", "sonata_paises_clean_AV.xlsx"))
sonata_countries
```

Remove duplicate:

```{r}
sonata_countries %>% count(ID) %>% filter(n > 1)
```

```{r}
sonata_countries %>% mutate(row_number = row_number()) %>% filter(ID == 2495)
```

```{r}
# Remove row number of the wrong entry
sonata_countries <- sonata_countries %>% slice(-619)
```

## Sensors DB

```{r}
sonata_sensors <-read_excel(
  here("data", "edited","to_merge", "sonata_sensor_data_clean.xlsx"),
  sheet = "sensors_AV")
sonata_sensors
```

Remove duplicate rows and combine rows:

```{r}
sonata_sensors <- sonata_sensors %>% 
  # Remove duplicate rows
  distinct() %>%
  # Unify Santinel-2 and Sentinel 2 to Sentinel-2
  mutate(Multispectral = if_else(Multispectral == "Santinel 2", "Sentinel-2", 
                                 Multispectral)) %>%
  # Combine all columns by ID, replace "" with NA
  group_by(ID) %>%
  summarise(across(everything(), ~ {
    vals <- sort(unique(na.omit(.)))
    if (length(vals) == 0) NA_character_ else paste(vals, collapse = "; ")
    }), .groups = "drop")
```

## Algorithms DB

```{r}
sonata_algorithms <-read_excel(
  here("data", "edited","to_merge", "algoritmos_inicial_corregido_OK_FINAL_JUNTO_AV.xlsx"))
sonata_algorithms
```

Remove duplicate:

```{r}
sonata_algorithms %>% count(ID) %>% filter(n > 1)
```

## Environmental data DB

```{r}
sonata_env <-read_csv(
  here("data", "edited","to_merge", "sonata_env_data_clean.csv"))
sonata_env
```

Modify to have the same format as other DBs. Columns Climate, Soil, Topography, Biotic. Each of them will (except Biotic) have the "main" variables separated by ;, variables that are not the main variables will be "Others", when variables were unspecified (i.e. people put 0 or 1), it will be "Unspecified". Biotic will have 0/1.

```{r}
sonata_env <- sonata_env %>%
  rowwise() %>%
  mutate(Clim_temp = if_else(Clim_temp == 1, "Temperature", NA),
         Clim_precip = if_else(Clim_precip  == 1, "Precipitation", NA),
         Clim_rad = if_else(Clim_rad == 1, "Radiation", NA),
         Clim_humid = if_else(Clim_humid == 1, "Humidiy", NA),
         Clim_wind = if_else(Clim_wind == 1, "Wind", NA),
         Clim_evap = if_else(Clim_evap == 1, "Evapotranspiration", NA),
         Clim_other = if_else(Clim_other == 1, "Other", NA),
         Clim_unspecif = if_else(Clim_unspecif == 1, "Unspecified", NA),
         Soil_type = if_else(Soil_type == 1, "Type", NA),
         Soil_text = if_else(Soil_text == 1, "Texture", NA),
         Soil_moist = if_else(Soil_moist == 1, "Moisture", NA),
         Soil_depth = if_else(Soil_depth == 1, "Depth", NA),
         Soil_ph = if_else(Soil_ph == 1, "pH", NA),
         Soil_carbon = if_else(Soil_carbon == 1, "Carbon", NA), 
         Soil_rough = if_else(Soil_rough == 1, "Roughness", NA),
         Soil_ferti = if_else(Soil_ferti == 1, "Fertility", NA),
         Soil_nitro = if_else(Soil_nitro == 1, "Nitrogen", NA),
         Soil_bulk = if_else(Soil_bulk == 1, "Bulk density", NA),
         Soil_other = if_else(Soil_other == 1, "Other", NA),
         Soil_unspecif = if_else(Soil_unspecif == 1, "Unspecified", NA),
         Topo_aspect =  if_else(Topo_aspect == 1, "Aspect", NA),
         Topo_slope = if_else(Topo_slope == 1, "Slope", NA),
         Topo_elev = if_else(Topo_elev == 1, "Elevation", NA),
         Topo_twi = if_else(Topo_twi == 1, "TWI", NA),
         Topo_curv = if_else(Topo_curv == 1, "Curvature", NA),
         Topo_other = if_else(Topo_other == 1, "Other", NA),
         Topo_unspecif = if_else(Topo_unspecif == 1, "Unspecified", NA)
         ) %>%
  ungroup() %>%
  unite("Climate", Clim_temp:Clim_unspecif, sep = "; ", na.rm = TRUE,
        remove = TRUE) %>%
  mutate(Climate = na_if(Climate, "")) %>%
  unite("Soil", Soil_type:Soil_unspecif, sep = "; ", na.rm = TRUE,
        remove = TRUE) %>%
  mutate(Soil = na_if(Soil, "")) %>%
  unite("Topography", Topo_aspect:Topo_unspecif, sep = "; ", na.rm = TRUE,
        remove = TRUE) %>%
  mutate(Topography = na_if(Topography, "")) %>%
  mutate(Biotic = Biotic_bin) %>%
  select(ID, Climate, Soil, Topography, Biotic)
```

# Merge

```{r}
sonata_db <- full_join(sonata_main, sonata_countries) %>%
  full_join(sonata_sensors) %>%
  full_join(sonata_algorithms) %>%
  full_join(sonata_env)
```

# Remove ID where I found errors

I found errors in these IDs when looking at the Excel files (e.g. the ID did not correspond to the reader and the info did not match that ID).

```{r}
sonata_db <- sonata_db %>% filter(ID != 66)
```

# Change errors in sensor

```{r}
sonata_db <- sonata_db %>%
  mutate(
    Multispectral = case_when(
      str_detect(Hyperspectral, "Landsat") & is.na(Multispectral) ~ "Landsat",
      str_detect(Hyperspectral, "Landsat") & !is.na(Multispectral) ~ 
        paste0(Multispectral, "; Landsat"),
      TRUE ~ Multispectral)
    ) %>%
  mutate(Hyperspectral = if_else(str_detect(Hyperspectral, "Landsat"),
                                 NA_character_, Hyperspectral))
```

# Order columns

```{r}
sonata_db <- sonata_db %>%
  select(ID, Reader, Year, Purpose, keyword1, keyword2, keyword3, `Study area`,
         Country_list, Extent, RGB, Multispectral, Hyperspectral, LiDAR, RaDAR,
         Thermal, Climate, Soil, Topography, Biotic, Visual, IndexClass, OBIA, 
         Regression, `Time series analysis`, ML, AI, Highlight)
```

# Save as csv

```{r}
write_csv(sonata_db, "data/clean/sonata_db_20250731.csv")
```

# Session info

```{r}
sessionInfo()
```
