---
title: "Anaylises for SONATA review"
subtitle: "Plots"
author: "Alicia Valdés"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
  html_notebook: default
---

# Load libraries

```{r}
library(here)
library(readxl)
library(tidyverse)
library(cowplot)
library(ggridges)
library(patchwork)
library(writexl)
```

# Read data

```{r}
sonata_db <- read_csv(
  here("data", "clean","sonata_db_20250731.csv")) 
sonata_db
```

# Create sensor and algorithm columns

```{r}
sonata_db <- sonata_db %>%
  rowwise() %>%
  mutate(Sensor = case_when(
    rowSums(!is.na(across(c(RGB, Multispectral, Hyperspectral, LiDAR, RaDAR, 
                            Thermal)))) == 0 ~ NA_character_,
    TRUE ~ paste(
      c(
        if (!is.na(RGB)) "RGB" else NULL,
        if (!is.na(Multispectral)) "Multispectral" else NULL,
        if (!is.na(Hyperspectral)) "Hyperspectral" else NULL,
        if (!is.na(LiDAR)) "LiDAR" else NULL,
        if (!is.na(RaDAR)) "Radar" else NULL,
        if (!is.na(Thermal)) "Thermal" else NULL
      ),
      collapse = "; "
    )
  )) %>%
  mutate(Algorithm = case_when(
    rowSums(!is.na(across(c(Visual, IndexClass, OBIA, Regression,
                            `Time series analysis`, ML, AI)))) == 0 ~
      NA_character_,
    TRUE ~ paste(
      c(
        if (!is.na(Visual)) "Visual" else NULL,
        if (!is.na(IndexClass)) "IndexClass" else NULL,
        if (!is.na(OBIA)) "OBIA" else NULL,
        if (!is.na(Regression)) "Regression" else NULL,
        if (!is.na(`Time series analysis`)) "Time series analysis" else NULL,
        if (!is.na(ML)) "ML" else NULL,
        if (!is.na(AI)) "AI" else NULL
      ),
      collapse = "; "
    )
  ))
```

# General

```{r}
sonata_db %>%
  mutate(Year_bin = paste0(
    1986 + 5 * ((Year - 1986) %/% 5),
    "–",
    1986 + 5 * ((Year - 1986) %/% 5) + 4
  )) %>%
  count(Year_bin) %>%
  mutate(
    Year_bin = factor(Year_bin, 
                      levels = unique(Year_bin[order(
                        as.numeric(sub("–.*", "", Year_bin)))])),
    n = as.numeric(n)
    ) %>%
  ggplot(aes(x = Year_bin, y = n, group = 1)) +
  geom_point(color = "azure4", size = 4) +
  geom_line(color = "azure4", size = 1) +
  theme_bw() +
  labs(x = "5-year interval", y = "Number of studies")
```
# Purpose per year

```{r}
sonata_db %>%
  count(Year, Purpose) %>%
  group_by(Year) %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  mutate(percentage = (n / total) * 100) %>%
  ggplot(aes(x = Year, y = n, fill = Purpose)) + 
  geom_area(position = "stack", color = "black", alpha = 0.8) +
  theme_bw() + ylab("Number of studies") +
  ggtitle("Purpose: area plot with n studies per year")
```

```{r}
sonata_db %>%
  ggplot(aes(x = Year, fill = Purpose)) +
  geom_density(alpha = 0.8, position = "stack", bw = 0.5) +
  labs(x = "Year", y = "Density", fill = "Purpose") +
  theme_bw() +
  ggtitle("Purpose: density plot per year")
```

```{r}
sonata_db %>%
  mutate(Year_bin = paste0(
    1986 + 5 * ((Year - 1986) %/% 5),
    "–",
    1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
  count(Year_bin, Purpose) %>%
  group_by(Year_bin) %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  mutate(percentage = (n / total) * 100) %>%
  ggplot(aes(x = Year_bin, y = n, fill = Purpose)) + 
  geom_col(position = "stack", color = "black", alpha = 0.8) +
  theme_bw() + ylab("Number of studies") + xlab("5-year interval") +
  ggtitle("Purpose: bar plot with n studies per 5-year interval")
```

```{r}
sonata_db %>%
  mutate(Year_bin = paste0(
    1986 + 5 * ((Year - 1986) %/% 5),
    "–",
    1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
  count(Year_bin, Purpose) %>%
  group_by(Year_bin) %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  mutate(percentage = (n / total) * 100) %>%
  ggplot(aes(x = Year_bin, y = percentage, fill = Purpose)) + 
  geom_col(position = "stack", color = "black", alpha = 0.8) +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position = position_stack(vjust = 0.5), size = 3, color = "black") +
  theme_bw() + ylab("% of studies") + xlab("5-year interval") +
  ggtitle("Purpose: bar plot with % studies per 5-year interval")
```

```{r}
sonata_db %>%
  mutate(
    Year_bin_start = 1986 + 5 * ((Year - 1986) %/% 5),
    Year_bin_mid = Year_bin_start + 2  # midpoint of 5-year interval
  ) %>%
  count(Year_bin_mid, Purpose) %>%
  group_by(Year_bin_mid) %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  mutate(percentage = (n / total) * 100) %>%
  ggplot(aes(x = Year_bin_mid, y = n, fill = Purpose)) + 
  geom_area(position = "stack", color = "black", alpha = 0.8) +
  theme_bw() +
  ylab("Number of studies") +
  xlab("Year (midpoint of 5-year intervals)") +
  ggtitle("Purpose: area plot with n studies per 5-year interval")
```

# Sensor per year

```{r}
sonata_db %>%
  mutate(RGB_bin = if_else(is.na(RGB), 0, 1),
         Multispectral_bin = if_else(is.na(Multispectral), 0, 1),
         Hyperspectral_bin = if_else(is.na(Hyperspectral), 0, 1),
         LiDAR_bin = if_else(is.na(LiDAR), 0, 1),
         RaDAR_bin = if_else(is.na(RaDAR), 0, 1),
         Thermal_bin = if_else(is.na(Thermal), 0, 1),
         ) %>%
  select(ID, Year, RGB_bin, Multispectral_bin, Hyperspectral_bin, LiDAR_bin,
         RaDAR_bin, Thermal_bin) %>%
  pivot_longer(cols = RGB_bin:Thermal_bin, names_to = "Sensor", values_to = "Used") %>%
  filter(Used == 1) %>%
  count(Year, Sensor) %>%
  mutate(Sensor = str_remove(Sensor, "_bin")) %>%
  ggplot( aes(x = Year, y = n, fill = Sensor)) +
  geom_area(alpha = 0.8, color = "black") +
  ylab("Number of studies") +  
  ggtitle("Sensor: area plot with n studies per year") +
  theme_minimal()
```

```{r}
sonata_db %>%
  separate_rows(Sensor, sep = ";\\s*") %>%
  mutate(Year_bin = paste0(
    1986 + 5 * ((Year - 1986) %/% 5),
    "–",
    1986 + 5 * ((Year - 1986) %/% 5) + 4
  )) %>%
  count(Year_bin, Sensor) %>%
  group_by(Year_bin) %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  mutate(percentage = (n / total) * 100) %>%
  ggplot(aes(x = Year_bin, y = n, fill = Sensor)) + 
  geom_col(position = "stack", color = "black", alpha = 0.8) +
  theme_bw() + ylab("Number of studies") + xlab("5-year interval") +
  ggtitle("Sensor: bar plot with n studies per 5-year interval")
```

```{r}
sonata_db %>%
  separate_rows(Sensor, sep = ";\\s*") %>%
  mutate(Year_bin = paste0(
    1986 + 5 * ((Year - 1986) %/% 5),
    "–",
    1986 + 5 * ((Year - 1986) %/% 5) + 4
  )) %>%
  count(Year_bin, Sensor) %>%
  group_by(Year_bin) %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  mutate(percentage = (n / total) * 100) %>%
  ggplot(aes(x = Year_bin, y = percentage, fill = Sensor)) + 
  geom_col(position = "stack", color = "black", alpha = 0.8) +
  theme_bw() + ylab("% of studies") + xlab("5-year interval") +
  ggtitle("Sensor: bar plot with % studies per 5-year interval")
```

```{r}
sonata_db %>%
  separate_rows(Sensor, sep = ";\\s*") %>%
  mutate(
    Year_bin_start = 1986 + 5 * ((Year - 1986) %/% 5),
    Year_bin_mid = Year_bin_start + 2  # midpoint of 5-year interval
  ) %>%
  count(Year_bin_mid, Sensor) %>%
  group_by(Year_bin_mid) %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  mutate(percentage = (n / total) * 100) %>%
  ggplot(aes(x = Year_bin_mid, y = n, fill = Sensor)) + 
  geom_area(position = "stack", color = "black", alpha = 0.8) +
  theme_bw() +
  ylab("Number of studies") +
  xlab("Year (midpoint of 5-year intervals)") +
  ggtitle("Sensor: area plot with n studies per 5-year interval")
```

# Algorithm per year

```{r}
sonata_db %>%
  separate_rows(Algorithm, sep = ";\\s*") %>%
  mutate(Year_bin = paste0(
    1986 + 5 * ((Year - 1986) %/% 5),
    "–",
    1986 + 5 * ((Year - 1986) %/% 5) + 4
  )) %>%
  count(Year_bin, Algorithm) %>%
  group_by(Year_bin) %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  mutate(percentage = (n / total) * 100) %>%
  ggplot(aes(x = Year_bin, y = percentage, fill = Algorithm)) + 
  geom_col(position = "stack", color = "black", alpha = 0.8) +
  theme_bw() + ylab("% of studies") + xlab("5-year interval") +
  ggtitle("Algorithm: bar plot with % studies per 5-year interval")
```

# Combined plots

## Bars

### Vertical

```{r}
combined_bars_vertical <- plot_grid(
  # N studies
  sonata_db %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin) %>%
    mutate(
      Year_bin = factor(Year_bin, 
                        levels = unique(Year_bin[order(
                          as.numeric(sub("–.*", "", Year_bin)))])),
      n = as.numeric(n)
    ) %>%
    ggplot(aes(x = Year_bin, y = n, group = 1)) +
    geom_point(color = "azure4", size = 2) +
    geom_line(color = "azure4") +
    theme_bw() +
    labs(x = NULL, y = "Number of studies") +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
          axis.title.x = element_blank()),
  # Purpose
  sonata_db %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin, Purpose) %>%
    group_by(Year_bin) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(percentage = (n / total) * 100,
           label = ifelse(percentage >= 5,
                          paste0(round(percentage), "%"), NA)) %>%
    ggplot(aes(x = Year_bin, y = percentage, fill = Purpose)) + 
    geom_col(position = "stack", color = "black", alpha = 0.8) +
    geom_text(aes(label = label),
              position = position_stack(vjust = 0.5), size = 3, color = "black") +
    theme_bw() + ylab("% of studies") + xlab(NULL) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
          axis.title.x = element_blank()),
  # Sensor
  sonata_db %>%
    # Keep only those that have sensor data
    filter(!is.na(Sensor))  %>%
    separate_rows(Sensor, sep = ";\\s*") %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin, Sensor) %>%
    group_by(Year_bin) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(percentage = (n / total) * 100,
           label = ifelse(percentage >= 5,
                          paste0(round(percentage), "%"), NA)) %>%
    ggplot(aes(x = Year_bin, y = percentage, fill = Sensor)) + 
    geom_col(position = "stack", color = "black", alpha = 0.8) +
    geom_text(aes(label = label),
              position = position_stack(vjust = 0.5), size = 3, color = "black") +
    theme_bw() + ylab("% of studies") + xlab(NULL) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
          axis.title.x = element_blank()),
  # Algorithm
  sonata_db %>%
    # Keep only those that have Algorithm data
    filter(!is.na(Algorithm))  %>%
    separate_rows(Algorithm, sep = ";\\s*") %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin, Algorithm) %>%
    group_by(Year_bin) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(percentage = (n / total) * 100,
           label = ifelse(percentage >= 5,
                          paste0(round(percentage), "%"), NA)) %>%
    ggplot(aes(x = Year_bin, y = percentage, fill = Algorithm)) + 
    geom_col(position = "stack", color = "black", alpha = 0.8) +
    geom_text(aes(label = label),
              position = position_stack(vjust = 0.5), size = 3, color = "black") +
    theme_bw() + ylab("% of studies") + xlab("5-year interval"),
  ncol = 1, align = "v", rel_heights = c(3, 7, 7, 8)
  )
```

```{r}
combined_bars_vertical
```

```{r}
ggsave(here("output", "figures", "combined_bars_vertical.tiff"), 
       plot = combined_bars_vertical, width = 20, height = 27, units = "cm", 
       dpi = 300)
ggsave(here("output", "figures", "combined_bars_vertical_200dpi.tiff"), 
       plot = combined_bars_vertical, width = 20, height = 27, units = "cm", 
       dpi = 200)
```

### Horizontal

```{r}
combined_bars_horizontal <- plot_grid(
  # N studies
  sonata_db %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin) %>%
    mutate(
      Year_bin = factor(Year_bin, 
                        levels = unique(Year_bin[order(
                          as.numeric(sub("–.*", "", Year_bin)), decrease = T)])),
      n = as.numeric(n)
    ) %>%
    ggplot(aes(x = n, y = Year_bin, group = 1)) +
    geom_point(color = "azure4", size = 3) +
    geom_line(color = "azure4", size = 1) +
    theme_bw() + 
    labs(x = "Number of studies", y = "5-year interval"),
  # Purpose
  sonata_db %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin, Purpose) %>%
    group_by(Year_bin) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(percentage = (n / total) * 100) %>%
    ggplot(aes(x = percentage, y = Year_bin, fill = Purpose)) + 
    geom_col(position = "stack", color = "black", alpha = 0.8) +
    geom_text(aes(label = paste0(round(percentage), "%")),
              position = position_stack(vjust = 0.5), size = 3, color = "black") +
    theme_bw() + ylab(NULL) + xlab("% of studies") +
    theme(legend.position = "top") +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), 
          axis.title.y = element_blank()),
  # Sensor
  sonata_db %>%
    separate_rows(Sensor, sep = ";\\s*") %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin, Sensor) %>%
    group_by(Year_bin) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(percentage = (n / total) * 100) %>%
    ggplot(aes(x = percentage, y = Year_bin, fill = Sensor)) + 
    geom_col(position = "stack", color = "black", alpha = 0.8) +
    geom_text(aes(label = paste0(round(percentage), "%")),
              position = position_stack(vjust = 0.5), size = 3, color = "black") +
    theme_bw() + ylab(NULL) + xlab("% of studies") +
    theme(legend.position = "top") +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), 
          axis.title.y = element_blank()),
  # Algorithm
  sonata_db %>%
    separate_rows(Algorithm, sep = ";\\s*") %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin, Algorithm) %>%
    group_by(Year_bin) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(percentage = (n / total) * 100) %>%
    ggplot(aes(x = percentage, y = Year_bin, fill = Algorithm)) + 
    geom_col(position = "stack", color = "black", alpha = 0.8) +
    geom_text(aes(label = paste0(round(percentage), "%")),
              position = position_stack(vjust = 0.5), size = 3, color = "black") +
    theme_bw() + ylab(NULL) + xlab("% of studies") +
    theme(legend.position = "top") +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), 
          axis.title.y = element_blank()),
  nrow = 1, align = "h", rel_widths = c(3, 7, 7, 7)
)
```

```{r}
combined_bars_horizontal
```

## Lines

```{r}
combined_lines <- plot_grid(
  # N studies
  sonata_db %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin) %>%
    mutate(
      Year_bin = factor(Year_bin, 
                        levels = unique(Year_bin[order(
                          as.numeric(sub("–.*", "", Year_bin)))])),
      n = as.numeric(n)
    ) %>%
    ggplot(aes(x = Year_bin, y = n, group = 1)) +
    geom_point(color = "azure4", size = 2) +
    geom_line(color = "azure4") +
    theme_bw() +
    labs(x = NULL, y = "Number of studies") +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
          axis.title.x = element_blank()),
  # Purpose
  sonata_db %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin, Purpose) %>%
    group_by(Year_bin) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(percentage = (n / total) * 100,
           label = ifelse(percentage >= 5,
                          paste0(round(percentage), "%"), NA)) %>%
    ggplot(aes(x = Year_bin, y = percentage, color = Purpose)) + 
    geom_line(aes(group = Purpose)) +
    geom_point(size = 2) +
    theme_bw() + ylab("% of studies") + xlab(NULL) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
          axis.title.x = element_blank()),
  # Sensor
  sonata_db %>%
    # Keep only those that have sensor data
    filter(!is.na(Sensor))  %>%
    separate_rows(Sensor, sep = ";\\s*") %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin, Sensor) %>%
    group_by(Year_bin) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(percentage = (n / total) * 100,
           label = ifelse(percentage >= 5,
                          paste0(round(percentage), "%"), NA)) %>%
    ggplot(aes(x = Year_bin, y = percentage, color = Sensor)) + 
    geom_line(aes(group = Sensor)) +
    geom_point(size = 2) +
    theme_bw() + ylab("% of studies") + xlab(NULL) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
          axis.title.x = element_blank()),
  # Algorithm
  sonata_db %>%
    # Keep only those that have Algorithm data
    filter(!is.na(Algorithm))  %>%
    separate_rows(Algorithm, sep = ";\\s*") %>%
    mutate(Year_bin = paste0(
      1986 + 5 * ((Year - 1986) %/% 5),
      "–",
      1986 + 5 * ((Year - 1986) %/% 5) + 4
    )) %>%
    count(Year_bin, Algorithm) %>%
    group_by(Year_bin) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(percentage = (n / total) * 100,
           label = ifelse(percentage >= 5,
                          paste0(round(percentage), "%"), NA)) %>%
    ggplot(aes(x = Year_bin, y = percentage, color = Algorithm)) + 
    geom_line(aes(group = Algorithm)) +
    geom_point(size = 2) +
    theme_bw() + ylab("% of studies") + xlab("5-year interval"),
  ncol = 1, align = "v", rel_heights = c(3, 7, 7, 8)
)
```

```{r}
combined_lines
```


```{r}
ggsave(here("output", "figures", "combined_lines.tiff"), 
       plot = combined_lines, width = 20, height = 27, units = "cm", 
       dpi = 300)
ggsave(here("output", "figures", "combined_lines_200dpi.tiff"), 
       plot = combined_lines, width = 20, height = 27, units = "cm", 
       dpi = 200)
```

# Sensor detailed

Recode some less abundant sensor subtypes to "Others".

```{r}
sonata_db_sensor_recoded <- sonata_db %>%
  mutate(RGB = str_replace_all(
    RGB,
    regex("Google Earth|Landsat TM|PhenoCam|QuickBird|Sentinel-2",
          ignore_case = TRUE), "Others"),
    Hyperspectral = str_replace_all(
      Hyperspectral,
      regex("AMSR-E|ASD Field Spectrometer|CASI|CHRIS-Proba|DESIS|EnMAP|GaoFen-5B (AHSI)|Kosmos-1939/MSU-E|MODIS|NEON AOP|PRISMA|SMOS|WindSat",
      ignore_case = TRUE), "Others"),
      LiDAR = str_replace_all(
        LiDAR,
        regex("Cloud|LVIS", ignore_case = TRUE), "Others"),
      RaDAR = str_replace_all(
        RaDAR,
        regex("Envisat ASAR|JERS-1 SAR|RADARSAT-2|TerraSAR-X",
              ignore_case = TRUE), "Others"),
      Thermal = str_replace_all(
        Thermal,
        regex("ASCAT|ASTER|CYGNSS|DMSP-OLS|LST|NPP-VIIRS|PAR sensor|Passive microwave|TRMM 3B42|UAV",
              ignore_case = TRUE),"Others")) %>%
  # Replace Landsat whatever by Landsat
  mutate(RGB = str_replace_all(
    RGB, 
    regex("Landsat TM|Landsat-4|Landsat-5|Landsat-7|Landsat-8|Landsat-9",
          ignore_case = TRUE), "Landsat"),
    Multispectral = str_replace_all(
      Multispectral, 
      regex("Landsat TM|Landsat-4|Landsat-5|Landsat-7|Landsat-8|Landsat-9", 
            ignore_case = TRUE), "Landsat"),
    Hyperspectral = str_replace_all(
      Hyperspectral, 
      regex("Landsat TM|Landsat-4|Landsat-5|Landsat-7|Landsat-8|Landsat-9",
            ignore_case = TRUE), "Landsat"),
    LiDAR = str_replace_all(
      LiDAR,
      regex("Landsat TM|Landsat-4|Landsat-5|Landsat-7|Landsat-8|Landsat-9",
            ignore_case = TRUE), "Landsat"),
    RaDAR = str_replace_all(
      RaDAR, regex("Landsat[^;]", ignore_case = TRUE), "Landsat"),
    Thermal = str_replace_all(
      Thermal, 
      regex("Landsat TM|Landsat-4|Landsat-5|Landsat-7|Landsat-8|Landsat-9",
            ignore_case = TRUE),"Landsat")) %>%
  # Correct "Others; Others"
  mutate(
    RGB = if_else(RGB == "Others; Others", "Others", RGB),
    Hyperspectral = if_else(Hyperspectral == "Others; Others", "Others",
                            Hyperspectral),
    LiDAR = if_else(LiDAR == "Others; Others", "Others", LiDAR),
    RaDAR = if_else(RaDAR == "Others; Others", "Others", RaDAR),
    Thermal = if_else(Thermal == "Others; Others", "Others", Thermal)
    ) %>%
  # Replace hyperspectral subtype that did not work before
  mutate(Hyperspectral = if_else(Hyperspectral == "GaoFen-5B (AHSI); Others",
                                 "Others", Hyperspectral))
```

## Per year

Info about sensor subtypes. Percentages calculated from IDs with info on sensor subtype.

```{r}
# Pivot longer to get all sensor categories in one column
sensor_long <- sonata_db_sensor_recoded %>%
  pivot_longer(cols = c(RGB, Multispectral, Hyperspectral, LiDAR, RaDAR, Thermal),
               names_to = "Sensor_type", values_to = "Sensor_subtype") %>%
  # Remove those without info on sensor subtype
  filter(!is.na(Sensor_subtype)) %>%
  separate_rows(Sensor_subtype, sep = ";\\s*") %>%
  group_by(Year, Sensor_type, Sensor_subtype) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Year, Sensor_type) %>%
  mutate(percentage = (Count / sum(Count)) * 100) %>%
  ungroup() %>%
  mutate(Sensor_type = factor(Sensor_type,
                              levels = c("RGB", "Multispectral",
                                         "Hyperspectral", "LiDAR", "RaDAR",
                                         "Thermal")))
```

```{r}
sensor_long %>%
  group_by(Sensor_type, Sensor_subtype) %>%
  summarise(total_counts = sum(Count)) %>%
  ungroup() %>%
  print(n = 100)
```

```{r}
# Dividir por tipo de sensor
sensor_groups <- sensor_long %>%
  group_split(Sensor_type)
```

### Bars

```{r}
sensor_bars <- map2(sensor_groups, seq_along(sensor_groups), ~ {
  p <- ggplot(.x, aes(x = Year, y = percentage, fill = Sensor_subtype)) +
    geom_col() +
    theme_bw() +
    labs(title = unique(.x$Sensor_type), y = "% of studies", x = if (.y == length(sensor_groups)) "Year" else NULL) +
    theme(
      legend.position = "right",
      legend.title = element_blank(),
      # Solo mostrar etiquetas del eje x en el último gráfico
      axis.title.x = if (.y != length(sensor_groups)) element_blank() else element_text(),
      axis.text.x = if (.y != length(sensor_groups)) element_blank() else element_text(),
      axis.ticks.x = if (.y != length(sensor_groups)) element_blank() else element_line()
    ) +
    guides(fill = guide_legend(ncol = 1))  # vertical legend
  
  return(p)
})

# Combine wtih patchwork
sensor_detail_bars <- wrap_plots(sensor_bars, ncol = 1)
sensor_detail_bars
```

### Lines

```{r}
sensor_lines <- map2(sensor_groups, seq_along(sensor_groups), ~ {
  p <- ggplot(.x, aes(x = Year, y = percentage, color = Sensor_subtype)) +
    geom_point() +
    geom_line() +
    theme_bw() +
    labs(title = unique(.x$Sensor_type), y = "% of studies") +
    theme(legend.position = "right", legend.title = element_blank()) +
    guides(color = guide_legend(ncol = 1))
  
  # Solo mostrar etiquetas del eje x en el último gráfico
  if (.y != length(sensor_groups)) {
    p <- p + theme(axis.title.x = element_blank(),
                   axis.text.x = element_blank(),
                   axis.ticks.x = element_blank())
  } else {
    p <- p + labs(x = "Year")
  }
  
  return(p)
})

# Combine with patchwork
sensor_detail_lines <- wrap_plots(sensor_lines, ncol = 1, guides = "collect", 
                            axis_titles = "collect")
sensor_detail_lines
```

```{r}
ggsave(here("output", "figures", "sensor_detail_lines.tiff"), 
       plot = sensor_detail_lines, width = 20, height = 27, units = "cm")
```

## 5-year bins

Info about sensor subtypes. Percentages calculated from IDs with info on sensor subtype.

```{r}
# Pivot longer to get all sensor categories in one column
sensor_long_5bins <- sonata_db_sensor_recoded %>%
  pivot_longer(cols = c(RGB, Multispectral, Hyperspectral, LiDAR, RaDAR, Thermal),
               names_to = "Sensor_type", values_to = "Sensor_subtype") %>%
  # Remove those without info on sensor subtype
  filter(!is.na(Sensor_subtype)) %>%
  separate_rows(Sensor_subtype, sep = ";\\s*") %>%
  mutate(Year_bin = paste0(1986 + 5 * ((Year - 1986) %/% 5), "–",
                           1986 + 5 * ((Year - 1986) %/% 5) + 4)) %>%
  group_by(Year_bin, Sensor_type, Sensor_subtype) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Year_bin, Sensor_type) %>%
  mutate(percentage = (Count / sum(Count)) * 100,
         label = ifelse(percentage >= 10,
                          paste0(round(percentage), "%"), NA)) %>%
  ungroup() %>%
  mutate(Sensor_type = factor(Sensor_type,
                              levels = c("RGB", "Multispectral",
                                         "Hyperspectral", "LiDAR", "RaDAR",
                                         "Thermal")))
```

```{r}
sensor_long_5bins %>%
  group_by(Sensor_type, Sensor_subtype) %>%
  summarise(total_counts = sum(Count)) %>%
  ungroup() %>%
  print(n = 100)
```

```{r}
# Dividir por tipo de sensor
sensor_groups_5bins <- sensor_long_5bins %>%
  group_split(Sensor_type)
```

### Bars

```{r}
sensor_bars_5bins <- map2(sensor_groups_5bins, seq_along(sensor_groups_5bins), 
                          ~ {
  p <- ggplot(.x, aes(x = Year_bin, y = percentage, fill = Sensor_subtype)) +
    geom_col(color = "black") +
    geom_text(aes(label = label),
              position = position_stack(vjust = 0.5),
              size = 3, color = "black") +
    theme_bw() +
    # Common x-axis
    scale_x_discrete(limits = 
                       sort(unique(unlist(lapply(sensor_groups_5bins, 
                                                 function(df) df$Year_bin))))) +
    labs(title = unique(.x$Sensor_type), y = "% of studies", x = if (.y == length(sensor_groups_5bins)) "Year" else NULL) +
    theme(
      legend.position = "right",
      legend.title = element_blank(),
      # Only show x-axis labels in the last graph
      axis.title.x = if (.y != length(sensor_groups_5bins)) 
        element_blank() else element_text(),
      axis.text.x = if (.y != length(sensor_groups_5bins)) 
        element_blank() else element_text(),
      axis.ticks.x = if (.y != length(sensor_groups_5bins)) 
        element_blank() else element_line()
    ) +
    guides(fill = guide_legend(ncol = 1))  # vertical legend
  
  return(p)
})


# Combine wtih patchwork
sensor_detail_bars_5bins <- wrap_plots(sensor_bars_5bins, ncol = 1)
sensor_detail_bars_5bins
```

```{r}
ggsave(here("output", "figures", "sensor_detail_bars.tiff"), 
       plot = sensor_detail_bars_5bins, width = 20, height = 27, units = "cm",
       dpi = 300)
ggsave(here("output", "figures", "sensor_detail_bars_200dpi.tiff"), 
       plot = sensor_detail_bars_5bins, width = 20, height = 27, units = "cm",
       dpi = 200)
```

### Lines

```{r}
sensor_lines_5bins <- map2(sensor_groups_5bins, seq_along(sensor_groups_5bins), 
                           ~ {
  p <- ggplot(.x, aes(x = Year_bin, y = percentage, color = Sensor_subtype)) +
    geom_line(aes(group = Sensor_subtype)) +
    geom_point() +
    theme_bw() +
    # Common x-axis
    scale_x_discrete(limits = 
                       sort(unique(unlist(lapply(sensor_groups_5bins, 
                                                 function(df) df$Year_bin))))) +
    labs(title = unique(.x$Sensor_type), y = "% of studies", x = if (.y == length(sensor_groups_5bins)) "Year" else NULL) +
    theme(
      legend.position = "right",
      legend.title = element_blank(),
      # Only show x-axis labels in the last graph
      axis.title.x = if (.y != length(sensor_groups_5bins)) 
        element_blank() else element_text(),
      axis.text.x = if (.y != length(sensor_groups_5bins)) 
        element_blank() else element_text(),
      axis.ticks.x = if (.y != length(sensor_groups_5bins)) 
        element_blank() else element_line()
    ) +
    guides(fill = guide_legend(ncol = 1))  # vertical legend
  
  return(p)
})


# Combine wtih patchwork
sensor_detail_lines_5bins <- wrap_plots(sensor_lines_5bins, ncol = 1)
sensor_detail_lines_5bins
```

```{r}
ggsave(here("output", "figures", "sensor_detail_lines.tiff"), 
       plot = sensor_detail_lines_5bins, width = 20, height = 27, units = "cm",
       dpi = 300)
ggsave(here("output", "figures", "sensor_detail_lines_200dpi.tiff"), 
       plot = sensor_detail_lines_5bins, width = 20, height = 27, units = "cm",
       dpi = 200)
```

# Algorithm detailed

Info about algorithm subtypes.

```{r}
# Pivot longer to get all sensor categories in one column
algorithm_types_subtypes <- sonata_db %>%
  pivot_longer(cols = c(Visual, IndexClass, OBIA, Regression, 
                        `Time series analysis`, ML, AI),
               names_to = "Algorithm_type", values_to = "Algorithm_subtype") %>%
  # Do not remove those without info on algorithm subtype
  separate_rows(Algorithm_subtype, sep = ";\\s*") %>%
  group_by(Algorithm_type, Algorithm_subtype) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Algorithm_type) %>%
  mutate(percentage = (Count / sum(Count)) * 100) %>%
  ungroup()
```

```{r}
write_xlsx(algorithm_types_subtypes, "data/edited/algorithm_types_subtypes.xlsx")
```

Recode some less abundant algorithm subtypes to "Others".

```{r}
sonata_db_algorithm_recoded <- sonata_db %>%
  # Only considering algorithm subtypes within AI IndexClass and ML
  mutate(AI = str_replace_all(
    AI,
    regex("Autoencoder|RNN-type network",
          ignore_case = TRUE), "Others"),
    IndexClass = str_replace_all(
      IndexClass,
      regex("SAR|Structural index",
      ignore_case = TRUE), "Others"),
    ML = str_replace_all(
        ML,
        regex("Feature selection|MaxEnt|PCA|Supervised classifier|Unsupervised classification|Unsupervised clustering", ignore_case = TRUE), "Others")) %>%
  # Correct "Others; Others"
  mutate(
    AI = if_else(AI == "Others; Others", "Others", AI),
    ML = if_else(ML == "Others; Others", "Others",
                            ML),
    IndexClass = if_else(IndexClass == "Others; Others", "Others", IndexClass)
    )
```

## 5-year bins

Info about algorithm subtypes. Percentages calculated from IDs with info on algorithm subtype.

```{r}
# Pivot longer to get all algorithm categories in one column
algorithm_long_5bins <- sonata_db_algorithm_recoded %>%
  pivot_longer(cols = c(Visual, IndexClass, OBIA, Regression, 
                        `Time series analysis`, ML, AI),
               names_to = "Algorithm_type", values_to = "Algorithm_subtype") %>%
  # Remove those without info on algorithm subtype
  filter(!is.na(Algorithm_subtype)) %>%
  separate_rows(Algorithm_subtype, sep = ";\\s*") %>%
  mutate(Year_bin = paste0(1986 + 5 * ((Year - 1986) %/% 5), "–",
                           1986 + 5 * ((Year - 1986) %/% 5) + 4)) %>%
  group_by(Year_bin, Algorithm_type, Algorithm_subtype) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Year_bin, Algorithm_type) %>%
  mutate(percentage = (Count / sum(Count)) * 100,
         label = ifelse(percentage >= 10,
                          paste0(round(percentage), "%"), NA)) %>%
  ungroup() %>%
  mutate(Algorithm_type = factor(Algorithm_type,
                              levels = c("Visual", "IndexClass", "OBIA",
                                         "Regression", "Time series analysis",
                                         "ML", "AI")))
```

```{r}
algorithm_long_5bins %>%
  group_by(Algorithm_type, Algorithm_subtype) %>%
  summarise(total_counts = sum(Count)) %>%
  ungroup() %>%
  print(n = 100)
```

```{r}
# Dividir por tipo de algorithm
algorithm_groups_5bins <- algorithm_long_5bins %>%
  # Only for IndexClass, ML, AI
  filter(Algorithm_type %in% c("IndexClass", "ML", "AI")) %>%
  group_split(Algorithm_type)
```

### Bars

```{r}
algorithm_bars_5bins <- map2(algorithm_groups_5bins,
                             seq_along(algorithm_groups_5bins), ~ {
  p <- ggplot(.x, aes(x = Year_bin, y = percentage, fill = Algorithm_subtype)) +
    geom_col(color = "black") +
    geom_text(aes(label = label),
              position = position_stack(vjust = 0.5),
              size = 3, color = "black") +
    theme_bw() +
    # Common x-axis
    scale_x_discrete(limits = 
                       sort(unique(unlist(lapply(algorithm_groups_5bins, 
                                                 function(df) df$Year_bin))))) +
    labs(title = unique(.x$Algorithm_type), y = "% of studies", x = if (.y == length(algorithm_groups_5bins)) "Year" else NULL) +
    theme(
      legend.position = "right",
      legend.title = element_blank(),
      # Only show x-axis labels in the last graph
      axis.title.x = if (.y != length(algorithm_groups_5bins)) 
        element_blank() else element_text(),
      axis.text.x = if (.y != length(algorithm_groups_5bins)) 
        element_blank() else element_text(),
      axis.ticks.x = if (.y != length(algorithm_groups_5bins)) 
        element_blank() else element_line()
    ) +
    guides(fill = guide_legend(ncol = 2))
  
  return(p)
})


# Combine wtih patchwork
algorithm_detail_bars_5bins <- wrap_plots(algorithm_bars_5bins, ncol = 1)
algorithm_detail_bars_5bins
```

```{r}
ggsave(here("output", "figures", "algorithm_detail_bars.tiff"), 
       plot = algorithm_detail_bars_5bins, width = 25, height = 14, units = "cm",
       dpi = 300)
ggsave(here("output", "figures", "algorithm_detail_bars_200dpi.tiff"), 
       plot = algorithm_detail_bars_5bins, width = 25, height = 14, units = "cm",
       dpi = 200)
```

### Lines

```{r}
algorithm_lines_5bins <- map2(algorithm_groups_5bins,
                             seq_along(algorithm_groups_5bins), ~ {
  p <- ggplot(.x, aes(x = Year_bin, y = percentage, color = Algorithm_subtype)) +
    geom_line(aes(group = Algorithm_subtype)) +
    geom_point() +
    theme_bw() +
    # Common x-axis
    scale_x_discrete(limits = 
                       sort(unique(unlist(lapply(algorithm_groups_5bins, 
                                                 function(df) df$Year_bin))))) +
    labs(title = unique(.x$Algorithm_type), y = "% of studies", x = if (.y == length(algorithm_groups_5bins)) "Year" else NULL) +
    theme(
      legend.position = "right",
      legend.title = element_blank(),
      # Only show x-axis labels in the last graph
      axis.title.x = if (.y != length(algorithm_groups_5bins)) 
        element_blank() else element_text(),
      axis.text.x = if (.y != length(algorithm_groups_5bins)) 
        element_blank() else element_text(),
      axis.ticks.x = if (.y != length(algorithm_groups_5bins)) 
        element_blank() else element_line()
    ) +
    guides(color = guide_legend(ncol = 2))
  
  return(p)
})


# Combine wtih patchwork
algorithm_detail_lines_5bins <- wrap_plots(algorithm_lines_5bins, ncol = 1)
algorithm_detail_lines_5bins
```

```{r}
ggsave(here("output", "figures", "algorithm_detail_lines.tiff"), 
       plot = algorithm_detail_lines_5bins, width = 25, height = 14, units = "cm",
       dpi = 300)
ggsave(here("output", "figures", "algorithm_detail_lines_200dpi.tiff"), 
       plot = algorithm_detail_lines_5bins, width = 25, height = 14, units = "cm",
       dpi = 200)
```

# Session info

```{r}
sessionInfo()
```
